---
title: "NYPD Shooting Analysis"
author: "JBrien Barcoma"
date: "2025-01-23"
output: 
      pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(forcats)
library(dplyr)
```

## Introduction

### Overview of Report

This report expounds the data analysis, visualization and insights derived from the dataset "NYPD Shooting Incident Data (Historic)" by the New York Police Department.

### Objectives of Analysis

Rather than simply exploring the dataset, we aim to purposely have the motivation to identify opportunities to reduce shooting incidents, in addition to obtaining any other insights that may assist or improve government operations in the topic of concern.

## Dataset Overview
### Data Source
The data was downloaded from the US Government's Open Data site (https://catalog.data.gov/dataset) on 1 January 2025. Incidents from January 2006 to December 2023

### Dataset Description
```{r}
df <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv")
```
The dateset consists of 21 columns and 28562 rows.

Below are the 21 column names and their respective data types.
```{r}
summary(df)
```



```{r}
generate_summary <- function(df) {
  # Create a summary data frame
  summary_df <- data.frame(
    Column_Name = names(df),                              # Column name
    Data_Type = I(sapply(df, class)),                    # Include all classes as a list
    Sample_Value = sapply(df, function(col) {
      if (length(na.omit(col)) > 0) na.omit(col)[1] else NA
    }), # First non-NA value
    NA_Count = sapply(df, function(col) sum(is.na(col))), # Number of missing values
    Unique_Values = sapply(df, function(col) length(unique(na.omit(col)))), # Unique non-NA values
    Min_Value = sapply(df, function(col) {
      if (is.numeric(col)) min(col, na.rm = TRUE) else NA
    }), # Min value for numeric columns
    Max_Value = sapply(df, function(col) {
      if (is.numeric(col)) max(col, na.rm = TRUE) else NA
    }), # Max value for numeric columns
    Mean_Value = sapply(df, function(col) {
      if (is.numeric(col)) mean(col, na.rm = TRUE) else NA
    }), # Mean value for numeric columns
    Mode_Value = sapply(df, function(col) {
      if (is.character(col) || is.factor(col)) {
        # Return the most frequent value for categorical columns
        tab <- table(col)
        names(tab)[which.max(tab)]
      } else {
        NA
      }
    }) # Mode for categorical columns
  )
  
  return(summary_df)
}

generate_summary(df)
```

We will remove INCIDENT_KEY, X_COORD_CD, Y_COORD_CD, and Lon_Lat. We will also rename the columns appropriately.

```{r}
df <- df %>%
  select(-c(
    INCIDENT_KEY,
    LOC_OF_OCCUR_DESC, 
    PRECINCT, 
    JURISDICTION_CODE, 
    LOC_CLASSFCTN_DESC,
    STATISTICAL_MURDER_FLAG,
    X_COORD_CD,
    Y_COORD_CD,
    Lon_Lat))
```

### Key Features and Variables

## Data Preprocessing
### Missing Values and Handling
Let us look at the percentage of missing values for each column, this will inform us on our interpretation of results in our later analysis.
```{r}
# 
```

### Data Cleaning & Transformation

## Dates

We will mutate the given date values and create new columns for Year, Month and Day of the Week and Hour.

```{r}
df <- df %>%
     mutate(OCCUR_DATE = mdy(OCCUR_DATE),
            OCCUR_TIME = hms(OCCUR_TIME))
```

```{r}
df <- df %>%
  mutate(date_column = dmy(OCCUR_DATE),
         month = month(OCCUR_DATE),
         year = year(OCCUR_DATE),
         day_of_week = wday(OCCUR_DATE, label = TRUE, abbr = FALSE),
         hour = hour(OCCUR_TIME))
```

## Exploratory Data Analysis
### Incidents by burough

This graph shows the count of incidents by burough
x <- Burough
y <- # of Incidents
```{r boro, echo=TRUE}
ggplot(df %>% filter(!is.na(LOCATION_DESC) & !is.null(LOCATION_DESC) & !LOCATION_DESC %in% c("(null)")), 
       aes(x = fct_infreq(LOCATION_DESC))) +
      geom_bar(fill = "purple") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      labs(title = "Bar Plot of Incidents by Location Description",
           x = "Location",
           y = "Number of Incidents")
```

### Incidents by Year by Burough

Let us look at the unique entries. We notice NA, null and NONE entries. We will remove them, but first we shall find out what percentage of incidents have NA, null and NONE entries.

```{r}
#list unique entries
unique(df$LOCATION_DESC)
```

```{r}
#convert null and none to NA
#create new column "isNA?"
#stacked bar chart of non-NA/NULL vs NA/NULL
```

```{r}
# Create new Month & Year Column
df <- df %>%
  mutate(date_column = dmy(date_column),
         month = month(date_column),
         year = year(date_column),
         day_of_week = wday(date_column, label = TRUE, abbr = FALSE))  # Extract full weekday name
```

Below is the chart for the number of incidents by location, for all years.

```{r pressure2, echo=TRUE}
ggplot(df %>% filter(!is.na(LOCATION_DESC) & !is.null(LOCATION_DESC)), aes(x = fct_infreq(LOCATION_DESC))) +
      geom_bar(fill = "purple") +
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
      labs(title = "Bar Plot of Incidents by Location Description",
           x = "Location",
           y = "Number of Incidents")
```

Let us look at the top 10 locations and break them down by year.

```{r}
```

### Incidents by Month by Year by Burough
```{r pressure3, echo=TRUE}
```

## Data Visualization & Insights

### Gender Ratio by Location
```{r pressure4, echo=TRUE}
```

### Map / Heat Map
```{r pressure5, echo=TRUE}
```

### Plot 3
```{r pressure6, echo=TRUE}
```

## Conclusion

```{r cars}
```

## References
### Session Info
```{r sessioninfo, echo=TRUE}
sessionInfo()
```
### Resources
